Accurate binding of sodium and calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-12-15 Manuscript/manuscript.tex

MD simulations were performed using the GROMACS \cite{Abraham15} simulation package (version 5.1.4). The simulation parameters used
in this work are summarized in Table~S1 in the SI. 
Compatibility with the openMM simulation package \cite{openmm7} was also
tested and is presented in the SI. 
The simulated trajectories and parameters files are available at \cite{??} \todo{To be uploaded to Zenodo}.

@@

The cationic surfactants were randomly mixed among the phospholipids to form bilayer structures with mole fractions of 10\%, 20\%, 30\%, 42\%, or 50\% of surfactant in the POPC bilayer. All these systems contained 50 POPC molecules per leaflet, 6340 water molecules, and 6, 14, 21, 35, or 50 surfactants per leaflet.  Chloride counter ions were used in all the simulations, while bromide was the counter ion in the experiment~\cite{scherer89}. For POPC, either Lipid14 with TIP3P~\cite{jorgensen83} or ECC-POPC with SPC/E~\cite{Berendsen1987} parameters  were used. The first 20~ns of the total simulation time of 200~ns was considered as an equilibration period and was thus omitted from the analysis. We checked that a sufficient lipid  exchange occurred during the simulations.
The trajectories and simulation files for unscaled systems are available from
Refs.~\citenum{POPClipid14T313K,POPClipid1410perCATsurfT313K,POPClipid1420perCATsurfT313K,POPClipid1430perCATsurfT313K,POPClipid1442perCATsurfT313K,POPClipid1450perCATsurfT313K}.

\todo{Provide the parameters for ECC-surfactant somewhere.
      Joe: They will appear along with the Zenodo upload of the trajectories.}

@@

\begin{figure}[tb!]
  \centering
  \includegraphics[width=8.2cm]{../Fig/ipython_nb/Order-parameters_form-factors_exp-L14-ECCL17_q80_sig89.pdf}
  \includegraphics[width=7.4cm]{../Fig/POPCstructure.eps}
  %\includegraphics[width=8.0cm]{../Fig/ipython_nb/Order-parameters_exp-L14-ECCL17_q80_sig89.pdf}
  %\includegraphics[height=8.4cm,angle=-90]{../Fig/form-f_exp-l14-eccl17.eps}
  \caption{\label{simVSexpNOions}
    Top: X-ray scattering form factors from simulations with the Lipid14 \cite{dickson14} and
    the ECC-POPC models compared with experiments~\cite{kucerka11} at 303~K.
    Middle: Order parameters of POPC head group, glycerol backbone and acyl chains 
    from simulations with the Lipid14 \cite{dickson14} and the ECC-POPC models
    compared with experiments \cite{ferreira13} at 300~K.
    The size of the markers for the head group order parameters correspond to
    the error estimate $\pm 0.02$ for experiments \cite{botan15,ollila16},
    while the error estimate for simulations is $\pm 0.005$.
    The size of the points for acyl chains are decreased by a factor of 3 to improve the clarity of the plot.
    Bottom: The chemical structure of POPC and the labeling of the carbon segments.
  } 
    \todo{Joe: I'm fine with the name change from ECC-lipids to ECC-POPC, but we have to be consistent throughout the text. I changed all accurences of ECC-lipids to ECC-POPC in the text, I yet need to change all figure legends then.}
\end{figure}

@@

\todo{Dynamics check is missing: MSD (Hector/Joe)}

@@

\begin{figure}[tb!]
  \centering
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/PN_angle_OrdPars-A-B_L14-ECCL17_q80_sig89_surf.pdf}
  \caption{\label{OrderParameterCHANGESsurf}
    The changes of head group order parameters and P-N vector orientation as a function of
    a molar fraction of the cationic surfactant dihexadecyldimethylammonium in a POPC bilayer
    from simulations and experiments \cite{scherer89} at 313 K.
  }
\end{figure}

@@

\begin{figure}[htb!]
  \centering
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/PN_angle_OrdPars-A-B-g3_L14-ECCL17_q80_sig89_CaCl.pdf}
  \caption{\label{fig:delta_ordPar_CaCl}
    Changes of the head group order parameters and P-N vector orientation of a POPC bilayer 
    as a function of the CaCl$_2$ concentration in bulk ($C_{ion}$)
    from simulations at 313 K together with experimental data 
    (DPPC (323\,K) \cite{akutsu81} and POPC (313\,K) \cite{altenbach84}). 
    The error estimate for bulk concentrations was approximately 10\,mM.
    Exception was the order of magnitude larger error in
    simulation with Lipid14 and ECC-ions due to the unconverged densities  (shown if Fig.~\ref{fig:cacl-dens}) because the 
    small simulation box. 
    Simulation data with Lipid14 and \AA{}qvist ion parameters at 298 K are taken directly from
    Refs.~\cite{lipid14POPC0mMNaClfiles,lipid14POPC350mMCaClfiles,lipid14POPC350mMCaClfilesNC}.
  }
  %Experimental values from \cite{akutsu81} for $g_3$ figure:
  %the value of $g_3$ order parameter of DPPC was -0.214 
  %in the absence of ions [average of two closely spaced splittings] and -0.211 in the presence
  %of 0.35 M CaCl$_2$ (at 59 Celcius). No effect of ions could be detected on DPPC bilayers labeled
  %at the C-2 segments of both fatty acyl chains.'' 
\end{figure}

@@

\begin{figure}[htbp!]
  \centering
  \includegraphics[height=19.5cm]{../Fig/ipython_nb/density_profiles_ca_cl_wat_phos_models-compar.pdf}
  \caption{\label{fig:cacl-dens}
    Number density profiles of \ce{Ca^{2+}}, \ce{Na^{+}} and \ce{Cl^-} along membrane normal axis
    for different force fields.
    In order to visualize the density profiles with a scale comparable to the profile of \ce{Ca^{2+}}, 
    the density profiles of~\ce{Cl^-} and \ce{Na^{+}} ions are divided by 2, and
    the density profiles of phosphate groups and water are divided by 5 and 200, respectively. 
    All simulations with \ce{CaCl2} shown here have the same molar concentration of ions in water ($C_{ion}'$=350~mM).
    The simulation with \ce{NaCl} has $C_{ion}'$=1000~mM.
    }
\end{figure}

@@

\begin{figure}[tb!]
  \centering
  \includegraphics[width=8.0cm]{../Fig/volmap_resid10_Ca_Cl_PO4Cent.png} %\\
%  \includegraphics[width=4.0cm]{../Fig/volmap_resid10_Ca_Cl_sn1Cent.png}
%  \includegraphics[width=3.6cm]{../Fig/volmap_resid10_Ca_Cl_sn2Cent.png}
  \caption{\label{fig:volmaps}
      Isocontours of probability density of \ce{Ca^{2+}} with respect to 
      the phosphate oxygens of POPC from ECC-POPC simulation.
      The probability density was evaluated around a single lipid, 
      after its strucutral alignment using only phosphate group. 
      %moiety, side chain 1 carbonyl group and side chain 2 carbonyl group.
      %Shown contours suggest that the dominant contribution 
      %to \ce{Ca^{2+}} binding comes from the phosphate oxygens,
      %whereas the interactions with any of the two carbonyl groups are considerably milder.
  }
  \todo{The figure should be updated as discussed last week.}
\end{figure}

@@

This work can be reached as a repository containing all data at \url{zenodo.org:\dots\dots\dots}.

>>>>

MD simulations were performed using the GROMACS \cite{Abraham15} simulation package (version 5.1.4). The simulation parameters used
in this work are summarized in Table~S1 in the SI. 
Compatibility with the openMM simulation package \cite{openmm7} was also
tested and is presented in the SI. 
The simulated trajectories and parameters files are available at
Refs.~\citenum{lipid14_POPC_Dang_and_ECC-ions_files, ECC-POPC_nacl_cacl2_files, ECC-POPC_nacl_cacl2_wat-models_files}.
\todo{To be uploaded to Zenodo, references are only placeholders. }.

@@

The cationic surfactants were randomly mixed among the phospholipids to form bilayer structures with mole fractions of 10\%, 20\%, 30\%, 42\%, or 50\% of surfactant in the POPC bilayer. All these systems contained 50 POPC molecules per leaflet, 6340 water molecules, and 6, 14, 21, 35, or 50 surfactants per leaflet.  Chloride counter ions were used in all the simulations, while bromide was the counter ion in the experiment~\cite{scherer89}. For POPC, either Lipid14 with TIP3P~\cite{jorgensen83} or ECC-POPC with SPC/E~\cite{Berendsen1987} parameters  were used. The first 20~ns of the total simulation time of 200~ns was considered as an equilibration period and was thus omitted from the analysis. We checked that a sufficient lipid  exchange occurred during the simulations.
The simulated trajectories and parameters files are available at Ref.~\citenum{ECC-POPC_cationic-surf_files} 
\todo{To be uploaded to Zenodo, references are only placeholders. }.
The trajectories and simulation files for unscaled systems are available at
Refs.~\citenum{POPClipid14T313K,POPClipid1410perCATsurfT313K,POPClipid1420perCATsurfT313K,POPClipid1430perCATsurfT313K,POPClipid1442perCATsurfT313K,POPClipid1450perCATsurfT313K}.

@@

\begin{figure}[tb!]
  \centering
  \includegraphics[width=8.2cm]{../Fig/ipython_nb/Order-parameters_form-factors_exp-L14-ECCL17_q80_sig89_POPC-struct.pdf}
  %\includegraphics[width=7.4cm]{../Fig/POPCstructure.eps}
  %\includegraphics[width=8.0cm]{../Fig/ipython_nb/Order-parameters_exp-L14-ECCL17_q80_sig89.pdf}
  %\includegraphics[height=8.4cm,angle=-90]{../Fig/form-f_exp-l14-eccl17.eps}
  \caption{\label{simVSexpNOions}
    Top: X-ray scattering form factors from simulations with the Lipid14 \cite{dickson14} and
    the ECC-POPC models compared with experiments~\cite{kucerka11} at 303~K. \\
    Middle: Order parameters of POPC head group, glycerol backbone and acyl chains 
    from simulations with the Lipid14 \cite{dickson14} and the ECC-POPC models
    compared with experiments \cite{ferreira13} at 300~K.
    The size of the markers for the head group order parameters correspond to
    the error estimate $\pm 0.02$ for experiments \cite{botan15,ollila16},
    while the error estimate for simulations is $\pm 0.005$.
    The size of the points for acyl chains are decreased by a factor of 3 to improve the clarity of the plot.
    Open/closed symbols are used for palmitoyl/oleoyl chains of POPC. \\
    Bottom: The chemical structure of POPC and the labeling of the carbon segments.
  } 
\end{figure}

@@

@@

\begin{figure}[tb!]
  \centering
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/PN_angle_OrdPars-A-B_L14-ECCL17_q80_sig89_surf.pdf}
  \caption{\label{OrderParameterCHANGESsurf}
    The changes of head group order parameters and P-N vector orientation as a function of
    a molar fraction of the cationic surfactant dihexadecyldimethylammonium in a POPC bilayer
    from simulations and experiments \cite{scherer89} at 313 K.
  }
  \todo{Joe: currenct figure depicts simulations with OPC3 water. Simulations with SPCE are running now.}
\end{figure}

@@

\begin{figure}[htb!]
  \centering
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/PN_angle_OrdPars-A-B-g3_L14-ECCL17_q80_sig89_CaCl.pdf}
  \caption{\label{fig:delta_ordPar_CaCl}
    Changes of the head group order parameters and P-N vector orientation of a POPC bilayer 
    as a function of the CaCl$_2$ concentration in bulk ($C_{ion}$)
    from simulations at 313 K together with experimental data 
    (DPPC (323\,K) \cite{akutsu81} and POPC (313\,K) \cite{altenbach84}). 
    The error estimate for bulk concentrations is approximately 10\,mM.
    The order of magnitude larger error in the
    simulation with Lipid14 and ECC-ions is due to unconverged bulk densities  (shown if Fig.~\ref{fig:cacl-dens}) limited by
    the simulation box. 
    Simulation data with Lipid14 and \AA{}qvist ion parameters at 298 K are taken directly from
    Refs.~\citenum{lipid14POPC0mMNaClfiles,lipid14POPC350mMCaClfiles,lipid14POPC350mMCaClfilesNC}.
  }
  %Experimental values from \cite{akutsu81} for $g_3$ figure:
  %the value of $g_3$ order parameter of DPPC was -0.214 
  %in the absence of ions [average of two closely spaced splittings] and -0.211 in the presence
  %of 0.35 M CaCl$_2$ (at 59 Celcius). No effect of ions could be detected on DPPC bilayers labeled
  %at the C-2 segments of both fatty acyl chains.'' 
\end{figure}

@@

\begin{figure}[htbp!]
  \centering
  \includegraphics[height=19.9cm]{../Fig/ipython_nb/density_profiles_ca_cl_wat_phos_models-compar.pdf}
  \caption{\label{fig:cacl-dens}
    Number density profiles of \ce{Ca^{2+}}, \ce{Na^{+}} and \ce{Cl^-} along membrane normal axis
    for different force fields.
    In order to visualize the density profiles with a scale comparable to the profile of \ce{Ca^{2+}}, 
    the density profiles of~\ce{Cl^-} and \ce{Na^{+}} ions are divided by 2, and
    the density profiles of phosphate groups and water are divided by 5 and 200, respectively. 
    All simulations with \ce{CaCl2} shown here have the same molar concentration of ions in water ($C_{ion}'$=350~mM).
    The simulation with \ce{NaCl} has $C_{ion}'$=1000~mM.
    }
\end{figure}

@@

\begin{figure}[tb!]
  \centering
  \includegraphics[width=8.0cm]{../Fig/isocontours_r37_ca_O-carb.png} %\\
%  \includegraphics[width=4.0cm]{../Fig/volmap_resid10_Ca_Cl_sn1Cent.png}
%  \includegraphics[width=3.6cm]{../Fig/volmap_resid10_Ca_Cl_sn2Cent.png}
  \caption{\label{fig:volmaps}
    Isocontours of spatial number density of \ce{Ca^{2+}} (dark blue, 0.0016~\AA$^{-3}$) 
    and POPC carbonyl oxygen atoms (light semi-transparent red, 0.007~\AA$^{-3}$, all POPC lipids contribute). 
    Calcium cations localize mostly around phosphate oxygens (red and bronze).
    Interactions with carbonyl oxygens is less likely than with phosphate oxygens, 
    and it is contributed more by other neighbouring phospholipids than by the same lipid. 
    Transparent structures are shown to depict the variability of choline configurations 
    (colour warps from red to blue along the simulation time). 
    The number density was evaluated for each lipid, 
    after its strucutral alignment using only phosphate group.
    MDAnalysis \cite{mdanalysis2011} library was used for 
    the calculations of the structural alignment and the spatial number density. 
    VMD \cite{hump96} was used for visualisation. 
    Carbon atoms are cyan, hydrogen atoms white, oxygen atoms red, nitrogen blue.
  }
 %\todo{The figure should be updated as discussed last week. Joe: Will only smoothen the isosurfaces now. }
\end{figure}

@@
